---
  title: "Understanding Fundamentalism"
  author: "Nikhil Raman"
  date: "December 27, 2017"
  output: 
    md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Aim

What are the factors that explain fundamentalism of an individual's religious belief? Using the General Social Survey dataset, this project attempts to identify some of the factors that account for the fundamentalism of a respondent's religious belief. 

## Data

### The following variable from the General Social Survey dataset was identified as the the dependent variable: 

**Fund**. This variable captures the fundamentalism of a respondent's religious belief on a 3 point scale. This variable was recoded into a binary variable indicating whether a respondent is a fundamentalist or not  

### The following variables from the General Social Survey dataset were identified as independent variables:

**Pray**. This variable captures how often a respondent prays on a 6 pointscale  

**Bible**. This variable captures how a respondent believes the bible should be viewed on a 4 pointscale  

**God**. This variable captures respondents' beliefs in the existence of god on a 6 point scale  

**Attend**. This varliable captures how often a respondent attends religious services on an 8 point scale   

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
# Importing relevant libraries...
library(magrittr)
library(dplyr)
library(readxl)
library(psych)
library(psych)
library(visreg)
```

```{r include=FALSE}
# Loading GSS data...
gss <- read.csv("/Users/antoshachekhonte/Documents/DataAnalysis/GSS.2006.csv.xls")
```

```{r include=FALSE}
# Reformatting selected variables...
gss$god <- as.numeric(ifelse(gss$god>6,NA,gss$god))
gss$fund <- as.numeric(ifelse(gss$fund==1,1,0))
gss$attend <- as.numeric(ifelse(gss$attend==9,NA,gss$attend))
gss$pray <- as.numeric(ifelse(gss$pray==1,6,
                   ifelse(gss$pray==2,5,
                          ifelse(gss$pray==3,4,
                                 ifelse(gss$pray==4,3,
                                        ifelse(gss$pray==5,2,
                                               ifelse(gss$pray==6,1,NA)))))))
gss$bible <- as.numeric(ifelse(gss$bible==1,4,
                    ifelse(gss$bible==2,3,
                           ifelse(gss$bible==3,2,
                                  ifelse(gss$bible==4,1,NA)))))
```

```{r include=FALSE}
# Subsetting the relevent variables...
vars <- c("god","fund","attend","pray","bible")
gss.sub <- gss[,vars]
```

## Simple Linear Model

```{r echo=FALSE}
# Building a preliminary model...
lm1 <- lm(fund~pray+god+attend+bible,gss.sub)
summary(lm1)
```

The independent variables *pray*, *attend*, *bible* and *god* account for **14.9%** of the variance in the dependent variable *fund*. All the independent variables are positively correlated with the dependent variable.  

*Pray* is positively correlated with *fund* and is associated with a coefficient of 0.0215, which means that a unit increase *pray* is associated with a 0.021 increase in *fund*, controlling for all other independent variables.  

Similarly, *attend* is associated with a coefficient of 0.0215, *bible* is associated with a coefficient of 0.126, and *god* is associated with a coefficient of 0.026.  

*Bible* has the strongest correlation with *fund*, meaning that the more uncritically a respondent accepts the bible as the word of god, the more likely he is to self report as a fundamentalist.  

## Binary Logistic Model

### Why should this model be used?

The linear probability model assumes that the error terms have the same variance. However, given that the dependent variable in this study is dichotomous, i.e., it can only take on two values, the above assumption is violated and heteroscedasticity is observed. Additionally, the linear probability model assumes the residuals must have a normal distribution. This assumption too, is violated, since residuals can only take one of two possible values.  

The logistic model overcomes these shortcomings by incorporating the logistic function (https://en.wikipedia.org/wiki/Logistic_function) in its definition.

```{r echo=FALSE}
logit1 <- glm(fund~pray+attend+bible+god,gss.sub,family=binomial)
summary(logit1)
```

From the above model summary, it can be observed that for every category increase in *pray*, the variable that measures a respondent's self-reported prayer frequency, the logit of *fund* increases by 0.16, net all other independent variables. Similarly, for a category increase in *attend*, the variable that measures the self reported religious attendance frequency, the logit of *fund* increases by 0.10. For a category increase in *bible*, the logit of *fund* increases by 0.73 and for a category increase in *god*, the logit of *fund* increases by 0.30.  

The two independent variables that have an especially strong relationship with the dependent variable are *bible*, which records the feelings of respondents about the bible, and *god*, which records the respondent's beliefs in god.

The independent variables *bible* and *god* have particularly strong relationships with *fund*. *Bible* takes on the following categorical values: "Other","Ancient Book", "Inspired Word" and "Actual Word". A category increase in *bible*, which represents an increase in the level at which a respondent accepts the precepts of the bible, is associated with an increase in the logit of self reporting to be a fundamentalist by 0.73. Similarly, a category increase in *god*, which represents an increase in the level at which a respondent accepts the existence of god, is associated with an increase in the logit of self reporting to be a fundamentalist by 0.31. 

### Odds-ratios from logit model

```{r echo=FALSE}
exp(coef(logit1))
```

The odds of being a fundamentalist increase by 17% with a category increase in the *pray*, net of all other independent variables. Similarly, the odds of being a fundamentalist increase by 10.5% with a category increase in the *attend* (net of all other variables). The odds of being a fundamentalist increase by 108% with a category increase in *bible* (net of all other variables), and by 36% with a category increase in *god* (net of all other variables).  
## Predicted Probabilities

### With pray, attend and god at the highest possible categories, what difference does moving from the lowest category of bible to the highest make?

```{r echo=FALSE}
predict(logit1,type="response",newdata=data.frame(pray=c(6,6),bible=c(1,4),attend=c(8,8),god=c(6,6)))
```

A movement from the lowest to the highest category of *bible*, with all other independent variables set to their highest category, is associated with an increase in the predicted probability of *fund* by 0.46 - a rather large increase!  

### With bible, attend and god at the highest possible categories, what difference does moving from the lowest category of pray to the highest make?  

```{r echo=FALSE}
predict(logit1, type = "response", newdata = data.frame(pray = c(1,6), bible = c(4,4), attend = c(8,8), god = c(6,6)))
```

A change from the lowest to the highest category of *pray*, with all other independent variables set at their highest categories, results in an increase in the predicted probability of *fund* by 0.19.  

This is interesting because it gives provides an intuition of the extent to which prayer frequency is related with self identification as a fundamentalist. The difference, on average, between the predicted *fund* scores of two individuals, one who responds to the question of "How often do you pray" with "Never", and the other who responds to the same with "Sevaral times a day" is 0.19.  

## Visual Study of Interaction between *Bible* and *God* 

```{r echo=FALSE}
# Changing datatype of bible simple and rerunning model...
gss.sub$biblesimple <- as.factor(gss.sub$bible)
logit2 <- glm(fund~pray+biblesimple+god+attend,gss.sub,family=binomial)
# Building visualization...
visreg(logit2,"god",by="biblesimple",partial=F,overlay=T,xlab="Belief in god",ylab="Predicted probability",scale="response",type="conditional",alpha=0.5)
```

For different answers to the question "Which statement comes closest to your view of the bible?", a change in the answer to the question "Do you believe in god" from "Don't believe", on one end, to "I believe without any doubt", on the other, results in different increases in the predicted probability of *fund* - for higher categories of *bible*, the increases in the predicted probability of *fund* with *god* are greater. 